---
jupyter:
  jupytext:
    text_representation:
      extension: .Rmd
      format_name: rmarkdown
      format_version: '1.2'
      jupytext_version: 1.19.1
  kernelspec:
    display_name: R
    language: R
    name: ir
---

# NYC Citibike Demand Analysis: Member vs Casual

```{r}
library(arrow)
library(dplyr)
library(ggplot2)
library(tidyr)
library(reshape2)
library(car)
library(bsts)
library(ggplot2)
```

# Data


## Dataset Overview

```{r}
bucket <- "citibike-pipeline-william-842549706992-us-east-1-an"

ds <- open_dataset(
  s3_bucket(bucket, region = "us-east-1")$path("processed"),
  format = "csv"
)
citybike.data <- ds |> collect()

glimpse(citybike.data)
```

```{r}
summary(citybike.data)
```

```{r}
table(citybike.data$weathercode)
```

```{r}
citybike.data <- citybike.data |>
  select(-yr, -mnth) |>
  mutate(
    season     = factor(season, levels = 1:4, labels = c("Winter", "Spring", "Summer", "Fall")),
    weekday    = factor(weekday, levels = 0:6),
    workingday = factor(workingday, levels = c(0, 1)),
    holiday    = factor(holiday, levels = c(0, 1)),
    weather_cat = case_when(
      weathercode %in% c(0, 1, 2)                ~ 1L,
      weathercode %in% c(3, 45)                  ~ 2L,
      weathercode %in% c(51, 53, 55, 61)         ~ 3L,
      weathercode %in% c(63, 65, 66, 80, 81, 82) ~ 4L,
      weathercode %in% c(71, 73, 75)             ~ 5L,
      weathercode == 95                           ~ 6L
    ) |> factor(levels = 1:6, labels = c("Clear", "Cloudy", "Light Rain", "Rain", "Snow", "Storm"))
  ) |>
  select(-weathercode)
```

```{r}
summary(citybike.data)
```

```{r}
table(citybike.data$weather_cat)
```

```{r}
citybike.na <- citybike.data |> 
  summarise(
    Total_Observations = n(),
    Missing_Values = sum(is.na(across(everything())))
  ) |>
  as.data.frame() 

print(citybike.na, row.names = FALSE)
```

# Exploratory Data Analysis


## Feature Distribution

```{r}
ggplot(citybike.data, aes(x = total_rides, fill = season)) +
  geom_histogram(bins = 30, alpha = 0.6, position = "identity") +
  labs(title = "Distribution of Daily Bike Rentals by Season",
       x = "Total Rides", y = "Frequency")
```

```{r}
citybike.data |>
  pivot_longer(cols = c(member_count, casual_count),
               names_to = "type", values_to = "count") |>
  ggplot(aes(x = workingday, y = count, fill = type)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Member vs Casual by Workingday",
       x = "Workingday", y = "Count")
```

```{r}
citybike.data |>
  pivot_longer(cols = c(member_count, casual_count),
               names_to = "type", values_to = "count") |>
  ggplot(aes(x = temp_max, y = count, color = type)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Member vs Casual by Temperature",
       x = "Max Temperature (°C)", y = "Count")
```

```{r}
citybike.data |>
  filter(precipitation < 20) |>
  pivot_longer(cols = c(member_count, casual_count),
               names_to = "type", values_to = "count") |>
  ggplot(aes(x = precipitation, y = count, color = type)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Member vs Casual by Precipitation (< 20mm)",
       x = "Precipitation (mm)", y = "Count")
```

```{r}
citybike.data |>
  pivot_longer(cols = c(member_count, casual_count),
               names_to = "type", values_to = "count") |>
  ggplot(aes(x = windspeed_max, y = count, color = type)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Member vs Casual by Wind Speed",
       x = "Max Wind Speed (km/h)", y = "Count")
```

## Feature Correlation

```{r}
numeric.features <- c("total_rides", "member_count", "casual_count",
                      "temp_max", "temp_min", "atemp_mean",
                      "precipitation", "windspeed_max", "humidity_mean")

cor.matrix <- cor(citybike.data[, numeric.features], use = "complete.obs")
cor.melted <- melt(cor.matrix)

ggplot(cor.melted, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile() +
  geom_text(aes(label = round(value, 2)), size = 3, color = "white") +
  scale_fill_gradient2(low = "grey90", mid = "steelblue", high = "navy",
                       midpoint = 0, limits = c(-1, 1)) +
  labs(title = "Correlation Matrix", x = "", y = "") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Multicollinearity Check (VIF)

```{r}
compute_vif <- function(data) {
  lm.proxy <- lm(total_rides ~ temp_max + precipitation + 
                   windspeed_max + humidity_mean + 
                   season + workingday + weather_cat, data = data)
  vif.result <- vif(lm.proxy)
  print(vif.result)
  invisible(vif.result)
}

compute_vif(citybike.data)
```

# Methodology


## Model Specification

```{r}
run_bsts <- function(data, target, formula, niter = 1000, seed = 42) {
  set.seed(seed)
  ss <- AddLocalLinearTrend(list(), data[[target]])
  ss <- AddSeasonal(ss, data[[target]], nseasons = 7)
  
  model <- bsts(formula,
                state.specification = ss,
                data = data,
                niter = niter)
  
  return(model)
}
```

```{r}
run_ols <- function(data, formula) {
  model <- lm(formula, data = data)
  return(model)
}
```

## Baseline


### Member Baseline

```{r}
member.baseline <- run_bsts(
  data    = citybike.data,
  target  = "member_count",
  formula = member_count ~ temp_max + precipitation + 
              windspeed_max + humidity_mean + 
              season + workingday + weather_cat
)
```

```{r}
plot(member.baseline, main = "Member Baseline BSTS")
```

```{r}
summary(member.baseline)
```

### Casual Baseline

```{r}
casual.baseline <- run_bsts(
  data    = citybike.data,
  target  = "casual_count",
  formula = casual_count ~ temp_max + precipitation + 
              windspeed_max + humidity_mean + 
              season + workingday + weather_cat
)
```

```{r}
plot(casual.baseline, main = "Casual Baseline BSTS")
```

```{r}
summary(casual.baseline)
```

## Interaction


### Member with Interaction

```{r}
member.interaction <- run_bsts(
  data    = citybike.data,
  target  = "member_count",
  formula = member_count ~ temp_max + precipitation + 
              windspeed_max + humidity_mean + 
              season + workingday + weather_cat +
              temp_max:workingday + precipitation:workingday
)
```

```{r}
plot(member.interaction, main = "Member BSTS with Interaction")
```

```{r}
summary(member.interaction)
```

### Casual with Interaction

```{r}
casual.interaction <- run_bsts(
  data    = citybike.data,
  target  = "casual_count",
  formula = casual_count ~ temp_max + precipitation + 
              windspeed_max + humidity_mean + 
              season + workingday + weather_cat +
              temp_max:workingday + precipitation:workingday
)
```

```{r}
plot(casual.interaction, main = "Casual BSTS with Interaction")
```

```{r}
summary(casual.interaction)
```

# Model Comparison


## Coefficient Comparison Across Models

```{r}
extract_coef <- function(model, model_name) {
  coef <- summary(model)$coefficients
  data.frame(
    term     = rownames(coef),
    mean     = coef[, "mean"],
    sd       = coef[, "sd"],
    inc.prob = coef[, "inc.prob"],
    model    = model_name
  )
}

coef.all <- rbind(
  extract_coef(member.baseline,    "Member Baseline"),
  extract_coef(casual.baseline,    "Casual Baseline"),
  extract_coef(member.interaction, "Member Interaction"),
  extract_coef(casual.interaction, "Casual Interaction")
)

coef.sig <- coef.all |> filter(inc.prob > 0.9)

ggplot(coef.sig, aes(x = term, y = mean, color = model)) +
  geom_point(position = position_dodge(0.5), size = 2) +
  geom_errorbar(aes(ymin = mean - 1.96 * sd, ymax = mean + 1.96 * sd),
                position = position_dodge(0.5), width = 0.3) +
  coord_flip() +
  labs(title = "Coefficient Comparison Across Models",
       x = "", y = "Coefficient Mean") +
  theme_minimal()
```

```{r}
coef.all |>
  filter(term == "workingday1", model %in% c("Member Baseline", "Casual Baseline")) |>
  select(model, mean) |>
  mutate(mean = round(mean, 2)) |>
  summarise(
    member_coef  = mean[model == "Member Baseline"],
    casual_coef  = mean[model == "Casual Baseline"],
    difference   = member_coef - casual_coef,
    ratio        = round(abs(member_coef / casual_coef), 2)
  ) |>
  print(row.names = FALSE)
```

## OLS Comparison 

```{r}
member.ols <- run_ols(
  data    = citybike.data,
  formula = member_count ~ temp_max + precipitation + 
              windspeed_max + humidity_mean + 
              season + workingday + weather_cat
)

casual.ols <- run_ols(
  data    = citybike.data,
  formula = casual_count ~ temp_max + precipitation + 
              windspeed_max + humidity_mean + 
              season + workingday + weather_cat
)

summary(member.ols)
summary(casual.ols)
```
